---
title: "Reproducible Research: Peer Assessment 1"
author: "David J. Weller-Fahy"
date: "February 05, 2016"
output: 
  html_document:
    keep_md: true
---

Clear the environment for those times the chunks are run out of knitr.  Comment the following line out if you do not want any existing environment variables removed.

```{r}
rm(list=ls())
```

If you want to keep the unzipped data file after the run, then set `KEEP` to `T`.

```{r}
KEEP = F
```

## Loading and preprocessing the data

Set the zip and data file names, and unzip the data file (if not already done).

```{r}
zfile = 'activity.zip'
cfile = 'activity.csv'
if (file.exists(zfile) & !file.exists(cfile))
  unzip(zfile, exdir = getwd())
```

Read in the data file and look at its structure.

```{r}
activity = read.csv(cfile)
str(activity)
```

The structure inspires me to perform a trick.

```{r}
cat(activity$steps[1:8], intToUtf8(c(66, 65, 84, 77, 65, 78, 33)))
```

Robin would be so proud.  Regardless, onward with the questions.

## What is mean total number of steps taken per day?



## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?

Clean up after ourselves.

```{r}
if (!KEEP) {
  unlink(cfile)
}
```
