---
title: "Reproducible Research: Peer Assessment 1"
author: "David J. Weller-Fahy"
date: "February 05, 2016"
output: 
  html_document:
    keep_md: true
---

Clear the environment for those times the chunks are run out of knitr.  Comment the following line out if you do not want any existing environment variables removed.

```{r}
rm(list=ls())
```

If you want to keep the unzipped data file after the run, then set `KEEP` to `T`.

```{r}
KEEP = F
```

## Loading and preprocessing the data

Set the zip and data file names, and unzip the data file (if not already done).

```{r}
zfile = 'activity.zip'
cfile = 'activity.csv'
if (file.exists(zfile) & !file.exists(cfile))
  unzip(zfile, exdir = getwd())
```

Read in the data file and look at its structure.

```{r}
activity = read.csv(cfile)
str(activity)
```

The structure inspires me to perform a trick.

```{r}
cat(activity$steps[1:8], intToUtf8(c(66, 65, 84, 77, 65, 78, 33)))
```

Robin would be so proud.  Regardless, onward with the questions.

## What is the mean total number of steps taken per day?

Sum the steps taken on each distinct date.  Note that some dates have no recorded steps: there are 61 distinct dates listed in the activity data, but only 53 have steps recorded for that date.

```{r}
daily = aggregate(steps ~ date, activity, sum, na.rm=T)
```

Display the distribution of the number of steps per day.  When considering how to label the histogram I found [this page][1] pretty useful.  I also (after creating the histogram for step 4) decided to keep the number of bins and y-axis limits the same for each, enabling easier comparison.

[1]: http://www.dummies.com/how-to/content/how-to-clearly-label-the-axes-on-a-statistical-his.html

```{r}
library(ggplot2)
library(scales)

histogram = function(df, var='steps', bins=nclass.FD(df[[var]]), title='Steps taken per day',
                     x.label='Steps', y.label='Number of days in each group of steps',
                     y.limits=c(0, 25)) {
  g = ggplot(df, aes_string(var)) +
    theme_bw() +
    geom_histogram(bins=bins) + #http://stats.stackexchange.com/a/862
    labs(title=paste(title, ' (n=', nrow(df), ')', sep='')) + 
    scale_y_continuous(breaks=pretty_breaks(), limits=y.limits) + #http://stackoverflow.com/a/19512397
    labs(y=y.label, x=x.label)
  g
}

histogram(daily)
```

In calculating the mean and median the `daily` variable is used, which does not contain any of the `NA` values present in the data.

```{r}
mean.steps.per.day = mean(daily$steps)
mean.steps.per.day

median.steps.per.day = median(daily$steps)
median.steps.per.day
```

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?

Clean up after ourselves.

```{r}
if (!KEEP) {
  unlink(cfile)
}
```
